Water discharge quantity has a variable impact on sediment transport capacity in subglacial channels

### 2023-09-21 Delaneyetal2023b.tex
figure update and formating for GRL

\documentclass[11pt]{article}
\usepackage [colon,square]{natbib}
\usepackage[mathlines]{lineno} %% <- mathlines turns on line numbering in equations
\usepackage{amsmath}
\usepackage{printlen}
\setlength{\parindent}{10pt}
\setlength{\parskip}{10pt}
\usepackage[explicit]{titlesec}
\usepackage{helvet}
\usepackage{array}
\usepackage{printlen}
\usepackage[a4paper, margin=1.3cm]{geometry}
\usepackage{color,hyperref}
\catcode`\_=11\relax
\newcommand\email[1]{\_email #1\q_nil}
\def\_email#1@#2\q_nil{%
  \href{mailto:#1@#2}{{\emailfont #1\emailampersat #2}}
}
\newcommand\emailfont{\sffamily}
\newcommand\emailampersat{{\color{red}\small@}}
\catcode`\_=8\relax

\usepackage{hyperref}
\usepackage[export]{adjustbox}
\usepackage{bibentry}
\renewcommand{\bibfont}{\footnotesize}
\usepackage{tikz}
\usepackage{setspace}
\usepackage[inline,shortlabels]{enumitem}
\usepackage[all]{nowidow}
\def\checkmark{\tikz\fill[scale=0.4](0,.35) -- (.25,0) -- (1,.7) -- (.25,.15) -- cycle;}
\pagenumbering{gobble}
\hypersetup{ % Remove ugly boxes around URLs
  colorlinks, linkcolor={black!100!black}, citecolor={black!50!black}, urlcolor={black!80!black}
}

\newcolumntype{L}[1]{>{\raggedright\arraybackslash}p{#1}}
\newcolumntype{C}[1]{>{\centering\arraybackslash\hspace{0pt}}p{#1}}

\linenumbers
\sloppy
\hbadness=99999

@@

\begin{document}
%% ------------------------------------------------------------------------ %%
% Title
%
% (A title should be specific, informative, and brief. Use
% abbreviations only if they are defined in the abstract. Titles that
% start with general keywords then specific terms are optimized in
% searches)
%
%% ------------------------------------------------------------------------ %%

% Example: \title{This is a test title}

\begin{center}
  \Large{\textbf{Water discharge quantity has a variable impact on sediment transport capacity in subglacial channels}}
  \normalsize

  Ian Delaney\footnote{Institut des dynamiques de la surface terrestre (IDYST), Universit\'{e} de Lausanne, B\^{a}timent G\'{e}opolis, 1015 Lausanne, Switzerland

    ianarburua.delaney@unil.ch},
  Andrew J. Tedstone\footnote{Department of Geosciences, University of Fribourg, Ch. du Musée 1700, Fribourg, Switzerland},
   Mauro A. Werder$^{3,4}$,
  Daniel Farinotti\footnote{Laboratory of Hydraulics, Hydrology and Glaciology (VAW), ETH-Z\"urich, H\"onggerbergring 26, 8093 89 Z\"urich, Switzerland}$^{,}$\footnote{Swiss Federal Institute for Forest, Snow and Landscape Research (WSL) Z\"uricherstrasse 111, 8903 1011 Birmensdorf, Switzerland}

\end{center}

\paragraph{Keypoints}
\begin{itemize}
\item Water discharge variations in subglacial channels are mainly accommodated by water velocity, not channel size, given their subglacial channel's evolution.

\item Greater variability in water velocity in subglacial channels causes sediment transport capacity to vary more than in subaerial channels.

\item In interpreting sediment transport records, water discharge may represent processes including proglacial sediment mobilization and sediment access driving variations in sediment discharge.
\end{itemize}

\begin{abstract}

  \noindent
  In both subglacial and subaerial channels, the sediment transport capacity for sediment size depends mainly on 1) the channel width over which the sediment is mobilized  and 2) water velocity, as this controls the shear stress that the flowing water applies to the channel bottom.
  In subaerial channels, changing water discharge  changes the water depth,  channel width, and velocity.
  However, in subglacial environments and on time scales of hours to weeks, changing water discharge  mainly alters the  water velocity, not channel area or width, because the subglacial channel maintains a largely fixed geometry.
  Here, parameterizations of sediment transport capacity for pressurized flow under glaciers and for subaerial open channel flow are applied to hydrographs from an Alpine glacier and the Greenland Ice Sheet.
  Results of these parameterizations show that sediment transport capacity varies far more in subglacial channels than subaerial ones over time and with respect to water discharge.
  Furthermore, high subglacial sediment transport capacity can persist across a wide range of water discharges.
  However, slowly evolving  channel width can accommodate some  variability in sediment transport capacity in subglacial channels.
  Despite the different responses of sediment transport capacity to water discharge in the two-channel types, measurements of sediment discharge from the catchments generally correlate better with subaerial hydraulics.
  This suggests that the sediment discharge records represent other processes such as sediment access or subaerial processes in the proglacial area.
  The different responses of water discharge on sediment transport capacity in subglacial channels, compared to subaerial ones, increase the occurrence of sediment transport capacity reaching mobilization thresholds, not necessarily related to water discharge quantity.

\end{abstract}

@@

Changes in glacier dynamics, geomorphology, and hydrology  have prompted numerous  recent studies of  sediment transport processes in cold regions \citep[e.g.][]{zhang2022}.
Increases in sediment transport  have been observed in Greenland \citep{bendixen2017}, the European Alps \citep{costa2017}, the Himalayas \citep{li2021}, and the Andes \citep{vergara2022}.
In some regions, increased water discharge and glacier melt have been interpreted to yield greater sediment transport capacity \citep{bendixen2017,costa2017,li2021}.
Observed changes to sediment transport in glacierized catchments require examining the processes controlling sediment discharge in these catchments and its variations with water discharge  \citep[e.g.][]{riihimaki2005,swift2005}.

Over long periods, processes such as glacier abrasion and quarrying sculpt landscapes and create sediment to be transported fluvially from under glaciers \citep[c.f.][]{hallet1979,iverson2012,ugelvig2018}.
On shorter time periods, pressurized water transports sediment from the under glacier \citep{walder1994,creyts2013,beaud2018,delaney2019}, should enough sediment be present subglacially (i.e. in a transport-limited regime).

In a transport-limited regime, sediment discharge responds to the sediment transport capacity defined as the amount of sediment the water can carry.
Sediment transport capacity depends on the shear stress between water and sediment it flows  over \citep{shields1936,meyer1948,engelund1967} and the width of the channel bottom $w_c$ over which to mobilize sediment.
The shear stress  across the channel width $\tau_t$ responds to the velocity of water $v$ flowing through the channel so that
\begin{linenomath*}
  \begin{equation}
    \label{eq:tau_t}
    \tau_t \propto w_c\, v^2,
  \end{equation}
\end{linenomath*}
%
where $w_c$ is the width of the channel. Following mass conservation, the velocity of the water flowing through a  channel is 
\begin{linenomath*}
  \begin{equation}
    \label{eq:v}
    v = \frac{Q}{S},
  \end{equation}
\end{linenomath*}
where $Q$ is water discharge,  and $S$ is the channel's wetted area. %\mauro{I think ``wetted area'' is better and less ambiguous, changed throughout.}
% \mauro{later $Q$ is used.  drop $_w$?}

In subaerial channels, operating with open channel flow, the wetted area $S$ of the channel evolves with changing water discharge $Q$, by changing both the channel's width and the water depth \citep{leopold1953}.  The change in water depth results in a commensurate increase in water velocity.  Thus,  fulfilling Equation~\ref {eq:v}, the higher discharge is accommodated by both an increase in wetted area and flow velocity. As a result, the shear stress across the channel $\tau_t$ increases according to Equation~\ref{eq:tau_t}.

The response of water velocity to changing water discharge in subglacial channels differs from subaerial ones.
The size of subglacial channels responds to the creep closure of the ice above and the opening of the channel by frictional heating of water flowing through the channel \citep{rothlisberger1972}.
As a result, the subglacial channel size only evolves relatively slowly over days to months, whereas water discharge can vary over hours \citep[e.g.][]{iken1986,andrews2014,nanni2020}.
Therefore, subglacial water flow behaves more like pipe flow, and changes in water discharge $Q$ are mainly accommodated with increased water velocity $v$ \citep[][]{alley1997}, because the size of the channel $S$ changes on a much longer timescale (Equation~\ref{eq:v} and Figure~\ref{fig:cartoon}).
It follows that an increase in water discharge $Q$ will cause the shear stress across the channel $\tau_t$ to respond with water velocity $v$, as opposed to subaerial channels where channel width $w_c$ evolves as well.

@@

As a result, sediment mobilization in subaerial and subglacial channels responds differently to changing water discharge.
These differences have been implicitly included in a wide range of available models that  quantify sediment transport from catchments in both subglacial and subaerial channels  \citep[e.g.][]{walder1994,alley1997,tucker1997,creyts2013,wickert2019,hewitt2019}.
The divergent response of changing water discharge on sediment mobilization capacity may well impact sediment dynamics at glacier margins where flow transitions from pressurized to open channel flow \citep[e.g.][]{lane2016,perolo2018}.
Additionally, the variable response of sediment transport capacity to water discharge in the two systems may affect the interpretation of sediment transport records in glacial systems \citep[e.g.][]{muller1968,richards2003,swift2005,ganti2016}.
Yet, little work has examined the differing relationship between variations in water discharge and sediment discharge in subglacial and subaerial channel \citep{alley1997}.
%Furthermore, thorough observations of water velocity and channel width in subglacial channels are limited and, to these authors' knowledge, not contrasted with subaerial conditions at the glacier margins. %\mauro{self-cite here?}

@@

Water and suspended sediment data are leveraged from an Alpine and an ice sheet setting.
The Alpine site (\alpine) is  Fieschergletscher in the Swiss Alps ($46^\circ\,29'\,07''$ N, $8^\circ\,08'\,3''$ E).
Water discharge and suspended sediment concentration were collected here at a $1$\unit{min} interval continuously over the period 2014--2021.
See \citet{felix2022} and \citet{felix2021} for more details.

The Leverett Glacier in Greenland (\icesheet) serves as the ice sheet setting.
Water discharge and suspended sediment concentration data were collected roughly $2$\unit{km} downstream from the terminus from June to August/September 2009 to 2012 ($67^\circ\,03'\,5''$ N, $50^\circ\,12'\,59''$ W).
This data is available from \citet{tedstone2017} at a $5$\,\unit{min} time interval of the 2012 season.
Note that in some cases the removal and relocation of sensors created variations in the sediment concentration not resulting from natural processes (see Supplement xxx).
% was presented in part in \citet{cowton2012} and

\section{Methods}
\label{sect:meth}
The parameterizations below (Sections ~\ref{sect:sub_mode}~and~\ref{sect:fluv}) represent relationships amongst water discharge, water velocity, and channel geometry in both subaerial and subglacial channels (Table \ref{table:vpm}).
Both parameterizations calculate   water velocity, shear stress, and width-integrated shear stress, upon which sediment transport depends \citep[Figure \ref{fig:cartoon}; ][]{shields1936}.
The evaluation of these variables omits the selection of a sediment transport relationship and grain-size parameter \citep[e.g.][]{shields1936,meyer1948} needed to calculate sediment transport capacity from shear stress.

\ian{values of f are still high}
\begin{table}[ht]
  \centering
  \caption{Variables, parameters, and constants used in this work. 
    Where two values are given, the first refers to  \alpine{}, a case from Fieschergletscher, and the second to a glacier marginal to the \icesheet{} a case from Leverett Glacier}
  \begin{tabular}{ l  c  c c }
    Name &Symbol&  Value&Units \\ \hline
    \textbf{Variables}  & & & \\
    Water discharge  & $Q$& & $\mathrm{m^{3}\,s^{-1}}$ \\
    Water velocity (glacier, subaerial)  & $v_g,\,v_{f}$& & $\mathrm{m\,s^{-1}}$ \\
    Channel wetted area (glacier, subaerial) &  $S_g, S_f$& & $\mathrm{m^2}$     \\
    Hydraulic diameter &$D_h$&&$\mathrm{m}$\\
    Width of channel floor (glacial, subaerial) & $w_{cg},w_{cf}$&  & $\mathrm{m}$     \\
    Hydraulic head &$\Delta h$&& $\mathrm{m}$\\
    Shear stress (glacier, subaerial) & $\tau_g,\,\tau_f$&& $\mathrm{Pa \, m^{-1}}$ \\
    Width-integrated shear stress (glacier, subaerial) & $\tau_{tg},\, \tau_{tf}$&& $\mathrm{Pa \, m^{-1}}$ \\

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% AGUJournalTemplate.tex: this template file is for articles formatted with LaTeX
%
% This file includes commands and instructions
% given in the order necessary to produce a final output that will
% satisfy AGU requirements, including customized APA reference formatting.
%
% You may copy this file and give it your
% article name, and enter your text.
%
% guidelines and troubleshooting are here:

%% To submit your paper:
\documentclass[draft]{agujournal2019}
\usepackage{url} %this package should fix any errors with URLs in refs.
\usepackage{lineno}
\usepackage[inline]{trackchanges} %for better track changes. finalnew option will compile document with changes incorporated.
\usepackage{soul}
\linenumbers
%%%%%%%
% As of 2018 we recommend use of the TrackChanges package to mark revisions.
% The trackchanges package adds five new LaTeX commands:
%
%  \note[editor]{The note}
%  \annote[editor]{Text to annotate}{The note}
%  \add[editor]{Text to add}
%  \remove[editor]{Text to remove}
%  \change[editor]{Text to remove}{Text to add}
%
% complete documentation is here: http://trackchanges.sourceforge.net/
%%%%%%%

\draftfalse

@@

\journalname{Geophysical Research Letters}

\begin{document}

\title{Water discharge quantity has a variable impact on sediment transport capacity in subglacial channels}

\authors{Ian Delaney\affil{1},  Daniel Farinotti\affil{2,3},  Andrew J. Tedstone\affil{4},   Mauro A. Werder\affil{2,3}}

\affiliation{1}{Institut des dynamiques de la surface terrestre (IDYST), Universit\'{e} de Lausanne, B\^{a}timent G\'{e}opolis, 1015 Lausanne, Switzerland}
\affiliation{2}{Laboratory of Hydraulics, Hydrology and Glaciology (VAW), ETH-Z\"urich, H\"onggerbergring 26, 8093 89 Z\"urich, Switzerland}
\affiliation{3}{Swiss Federal Institute for Forest, Snow and Landscape Research (WSL) Z\"uricherstrasse 111, 8903 1011 Birmensdorf, Switzerland}
\affiliation{4}{Department of Geosciences, University of Fribourg, Ch. du Musée 1700, Fribourg, Switzerland}

\correspondingauthor{Ian Delaney}{ianarburua.delaney@unil.ch}

\begin{keypoints}
\item Water discharge variations in subglacial channels are mainly accommodated by water velocity, not channel size, given their subglacial channel's evolution.
\item Greater variability in water velocity in subglacial channels causes sediment transport capacity to vary more than in subaerial channels.
\item In interpreting sediment transport records, water discharge may represent processes including proglacial sediment mobilization and sediment access driving variations in sediment discharge.
\end{keypoints}

\begin{abstract}
  Sediment transport in subglacial and subaerial channels depends on  water velocity and channel width, which controls the shear stress exerted on the channel bottom and the space over which to integrate it.
  In subaerial channels, depth, width, and velocity evolve quickly with water discharge, while in subglacial channels, water discharge changes mainly affect water velocity due to slow-evolving channel geometry.
  Parameterizations applied to hydrographs from an Alpine glacier and the Greenland Ice Sheet reveal greater variability in sediment transport capacity in subglacial channels.
  Furthermore, results show that high subglacial sediment transport rates can persist across various water discharges.
  Slowly evolving channel width accommodates some variability in subglacial channels.
  Despite differences in response to water discharge, sediment discharge measurements generally correlate better with subaerial hydraulics.
  This implies that other factors like sediment access or proglacial processes might be at play, affecting sediment transport capacity differently in subglacial and subaerial channels.
\end{abstract}

@@

Changes in glacier dynamics, geomorphology, and hydrology  have prompted numerous  recent studies of  sediment transport processes in cold regions \cite<e.g.>{zhang2022}.
Increases in sediment transport  have been observed in Greenland \cite{bendixen2017}, the European Alps \cite{costa2017}, the Himalayas \cite{li2021}, and the Andes \cite{vergara2022}.
In some regions, increased water discharge and glacier melt have been interpreted to yield greater sediment transport capacity \cite{bendixen2017,costa2017,li2021}.
Observed changes to sediment transport in glacierized catchments require examining the processes controlling sediment discharge in these catchments and its variations with water discharge  \cite<e.g.>{riihimaki2005,swift2005}.

Over long periods, processes such as glacier abrasion and quarrying sculpt landscapes and create sediment to be transported fluvially from under glaciers \cite<c.f.>{hallet1979,iverson2012,ugelvig2018}.
On shorter time periods, pressurized water transports sediment from the under glacier \cite{walder1994,creyts2013,beaud2018,delaney2019}, should enough sediment be present subglacially (i.e. in a transport-limited regime).

In a transport-limited regime, sediment discharge responds to the sediment transport capacity defined as the amount of sediment the water can carry.
Sediment transport capacity depends on the shear stress between water and sediment it flows  over \cite{shields1936,meyer1948,engelund1967} and the width of the channel bottom $w_c$ over which to mobilize sediment.
The shear stress  across the channel width $\tau_t$ responds to the velocity of water $v$ flowing through the channel so that
\begin{linenomath*}
  \begin{equation}
    \label{eq:tau_t}
    \tau_t \propto w_c\, v^2,
  \end{equation}
\end{linenomath*}
%
where $w_c$ is the width of the channel. Following mass conservation, the velocity of the water flowing through a  channel is 
\begin{linenomath*}
  \begin{equation}
    \label{eq:v}
    v = \frac{Q}{S},
  \end{equation}
\end{linenomath*}
where $Q$ is water discharge,  and $S$ is the channel's wetted area.

In subaerial channels, operating with open channel flow, the wetted area $S$ of the channel evolves with changing water discharge $Q$, by changing both the channel's width and the water depth \cite{leopold1953}.  The change in water depth results in a commensurate increase in water velocity.  Thus,  fulfilling Equation~\ref {eq:v}, the higher discharge is accommodated by both an increase in wetted area and flow velocity. As a result, the shear stress across the channel $\tau_t$ increases according to Equation~\ref{eq:tau_t}.

The response of water velocity to changing water discharge in subglacial channels differs from subaerial ones.
The size of subglacial channels responds to the creep closure of the ice above and the opening of the channel by frictional heating of water flowing through the channel \cite{rothlisberger1972}.
As a result, the subglacial channel size only evolves relatively slowly over days to months, whereas water discharge can vary over hours \cite<e.g.>{iken1986,andrews2014,nanni2020}.
Therefore, subglacial water flow behaves more like pipe flow, and changes in water discharge $Q$ are mainly accommodated with increased water velocity $v$ \cite{alley1997}, because the size of the channel $S$ changes on a much longer timescale (Equation~\ref{eq:v} and Figure~\ref{fig:cartoon}).
It follows that an increase in water discharge $Q$ will cause the shear stress across the channel $\tau_t$ to respond with water velocity $v$, as opposed to subaerial channels where channel width $w_c$ evolves as well.

@@

As a result, sediment mobilization in subaerial and subglacial channels responds differently to changing water discharge.
These differences have been implicitly included in a wide range of available models that  quantify sediment transport from catchments in both subglacial and subaerial channels  \cite<e.g.>{walder1994,alley1997,tucker1997,creyts2013,delaney2019,hewitt2019,wickert2019}.
The divergent response of changing water discharge on sediment mobilization capacity may well impact sediment dynamics at glacier margins where flow transitions from pressurized to open channel flow \cite<e.g.>{lane2016,perolo2018}.
Additionally, the variable response of sediment transport capacity to water discharge in the two systems may affect the interpretation of sediment transport records in glacial systems \cite<e.g.>{muller1968,richards2003,swift2005,ganti2016}.
Yet, little work has examined the differing relationship between variations in water discharge and sediment discharge in subglacial and subaerial channel \cite{alley1997}.
%Furthermore, thorough observations of water velocity and channel width in subglacial channels are limited and, to these authors' knowledge, not contrasted with subaerial conditions at the glacier margins. %\mauro{self-cite here?}

@@

Water and suspended sediment data are leveraged from an Alpine and an ice sheet setting.
The Alpine site (\alpine) is  Fieschergletscher in the Swiss Alps ($46^\circ\,29'\,07''$ N, $8^\circ\,08'\,3''$ E).
Water discharge and suspended sediment concentration were collected here at a $1$\unit{min} interval continuously over the period 2014--2021.
See \citeA{felix2022} and \citeA{felix2021} for more details.

The Leverett Glacier in Greenland (\icesheet) serves as the ice sheet setting.
Water discharge and suspended sediment concentration data were collected roughly $2$\unit{km} downstream from the terminus from June to August/September 2009 to 2012 ($67^\circ\,03'\,5''$ N, $50^\circ\,12'\,59''$ W).
Data is used from \citeA{tedstone2017} and \citeA{hawkings2014} at a $5$\,\unit{min} time interval from May 28, 2012 to August 8, 2012.
Two gaps in suspended sediment data of roughly day in length are interpolated across with a linear spline.

\section{Methods}
\label{sect:meth}
The parameterizations below (Sections ~\ref{sect:sub_mode}~and~\ref{sect:fluv}) represent relationships amongst water discharge, water velocity, and channel geometry in both subaerial and subglacial channels (Table \ref{table:vpm}).
Both parameterizations calculate   water velocity, shear stress, and width-integrated shear stress, upon which sediment transport depends \cite<Figure \ref{fig:cartoon}; >{shields1936}.
The evaluation of these variables omits the selection of a sediment transport relationship and grain-size parameter \cite<e.g.>{shields1936,meyer1948} needed to calculate sediment transport capacity from shear stress.

To evaluate the shear stress of water flowing across sediments underneath a glacier, the subglacial channel parameterization takes into account the channel geometry and the velocity of the flowing water.
To do this, a modified version of the lumped hydraulics model presented in \citeA{clarke1996} and \citeA{werder2010} is used.

Here, it is assumed that the water is transported through a subglacial channel flowing down the glacier; \cite<Figure~\ref{fig:cartoon}; >{rothlisberger1972}, and that the channel  size responds to frictional heating from water flow and creep closure by ice, using the Darcy-Weisbach formulation for water-flow through a pipe  \cite<e.g.>{rothlisberger1972,clarke2003}.
The formulation here does not consider the englacial storage of water.
Thus, it is assumed that the changing hydraulic head at the top of the glacier is negligible in the channel size evolution.
The evolution of subglacial channel size $S_g$ is given as
\begin{linenomath*}
  \begin{equation}
    \label{eq:dS_dt}
    \frac{\partial S_g}{\partial t} = C_1 \frac{Q \Delta h}{l} - C_2 \left(h_{o}-\frac{\Delta h}{2}\right)^n\,S_g,
  \end{equation}
\end{linenomath*}
\noindent where $C_1= (1-\rho_wc_pc_t)\,\frac{\rho_wg}{\rho_iL}$ and $C_2=2A(\frac{\rho_wg}{n})^n$ are constants (Table \ref{table:vpm}), $l$ is the length of subglacial channel, $Q$ is water discharge, $\Delta h$ is the hydraulic potential change from the glacier terminus to its top, $h_{o}= \frac{\rho_i}{\rho_w} h_{ice}$ is the ice overburden pressure, and $n$ is Glen's n \cite{glen1955} (usually $n=3$).
The first term on the right side of the equation represents the opening of the channel through frictional heating, while the following term represents the creep closure of the channel from ice deformation.

Following the Darcy-Weisbach, head drop $\Delta h$ is,
\begin{linenomath*}
  \begin{equation}
    \label{eq:dh}
    \Delta h \,  = l \,\frac{1}{g}\,s\,f_r\,\frac{Q^2}{D_h^5}
  \end{equation}
\end{linenomath*}
\noindent where $f_r$ is a friction factor, $D_h$ is the hydraulic diameter, and $l$ is the length overwhich head change occurs. $s$ is a factor accounting for channel geometry \cite{hooke1990}, calculated as
\begin{equation}
  \label{eq:Hf}
  s = \frac{2\,(\beta -\sin \beta)^2}{(\frac{\beta}{2}\,+\,\sin \frac{\beta}{2})^4},
\end{equation}
where $\beta$ is the central angle of the circular segment that comprises the channel (the so-called Hooke angle). Note that $\beta =\pi$ corresponds to a semi-circle and
smaller values of $\beta$ result in shallow, wide channels.
The hydraulic diameter $D_h$ is converted to wetted area $S_g$ given a channel geometry following \cite{hooke1990}
\begin{equation}
  \label{eq:Dh2S}
  S_g= \frac{D_h^2}{2}\,\frac{(\frac{\beta}{2}+\sin \frac{\beta}{2})^2}{\beta - \sin \beta},
\end{equation}
where $\beta$ is the central angle of the circular segment that comprises the channel (the so-called Hooke angle).

To parameterize the shear stress of water flowing across sediments in the subaerial channel,  the hydraulics parameterization presented in \citeA{tucker1997} is implemented.
Using mass conservation and the Darcy-Weisbach relationship as well as assuming that the channel is wide so that the hydraulic radius is well approximated by the flow depth, then
the shear stress $\tau_f$ at the river bed is
\begin{linenomath*}
  \begin{equation}
    \label{eq:DW_tau}
    \tau_f=\frac{\rho_w\,g^{\frac{2}{3}}\,f_p^{\frac{1}{3}}}{2}\, \Big(\frac{Q}{w_{cf}} \Big)^{\frac{2}{3}} \,\nabla z_c^{\frac{2}{3}},
  \end{equation}
\end{linenomath*}
where $\nabla z_c$ is the channel slope, and $f_p$ is the friction factor for subaerial channels.
Channel width $w_{cf}$ is
\begin{equation}
  \label{eq:wcf}
  w_{cf} = k \, Q^\frac{1}{2},
\end{equation}
%
where $k$ is a constant and the exponent is commonly given as $\frac{1}{2}$ \cite{leopold1953}.
%\mauro{This is for so called ``regime channels''}
Water velocity, $v_f$, is given by rearranging Equation \ref{eq:tau} as
\begin{equation}
  \label{eq:vf}
  v_f = \sqrt{\frac{8\,\tau_f}{f_p\,\rho_w}}.
\end{equation}
%
As above, the width-integrated shear stress is
\begin{equation}
    \label{eq:tautf}
    \tau_{tf}=w_{cf}\,\tau_f.
  \end{equation}

\subsection{Implementation}
\label{sect:imp}

The parameterizations are first applied to a reference test case for each glacier that assumes  subglacial channel with $\beta=\frac{\pi}{2}$ and friction factors $f_r$ and $f_p$ are tuned so that the parameterizations reproduce reasonable velocities for both  the Greenland Ice Sheet and Alpine glaciers \cite<$\sim\,1.6\,$\unit{m}$^3$\,\unit{s}$^{-1}$>{werder2010b,chandler2013}.
In both subaerial channels, the bed slope, $\nabla z_c$ is $0.05$.
Water velocity, shear stress, and width-integrated shear stress in the subglacial and subaerial channels from this formulation are presented below.

Second, to characterize the variability in sediment discharge capacity in subglacial compared to subaerial channels further, $100$  parameterizations are applied to a range of different channel geometry factors ($\beta$ and $k$) and friction factors ($f_r$ and $f_p$).
Runs with parameter combinations are culled if their mean subglacial water velocity over the season lies outside of the range between $0.75$\,\unit{m}$^3$\,\unit{s}$^{-1}$ and $2$\,\unit{m}$^3$\,\unit{s}$^{-1}$ or if subaerial water velocity lies outside of the range $0.3$\,\unit{m}$^3$\,\unit{s}$^{-1}$ and $1.2$\,\unit{m}$^3$\,\unit{s}$^{-1}$, in accordance with some measurements \cite<e.g.>{werder2010b,magnusson2012,chandler2013}.
The standard deviation of an output variable over the course of the season is used to quantify the variability of that output.

\subsection{Role of water discharge in sediment transport capacity}
\begin{figure}[h]
  \centering
  \includegraphics[width=0.9\linewidth]{Fig2.pdf}
  \caption{Parameterization outputs resulting from the hydrographs shown in panels a and e for the cases called \alpine (a-d) and \icesheet{} (e-h). Gray (black) lines in a and e represent the subglacial channel size (water discharge).  In other panels, blue lines represent outputs from the subglacial channel, while orange lines represent the subaerial channel.
    Data in plots are at $15$\unit{min} intervals.
    Arrows denote some events where the variable peaks in the subglacial channel prior to the subaerial one.
    Insets in e-h show the  peak melt event \icesheet{} denoted by the shaded area.
    Note that the x and y axes are different for \alpine{} and \icesheet{}.
      }
  \label{fig:model_outs}
\end{figure}
Following the parameterizations defined in Sections~\ref{sect:sub_mode}~and~\ref{sect:fluv}, the water velocity, shear stress, and width-integrated shear stress exhibit different seasonal evolutions and peaks. Water velocity and shear stress represent the potential for sediment mobilization, and width-integrated shear stress represents a proxy for the total sediment transport capacity across the channel bed (Figure~\ref{fig:model_outs}).

\begin{figure}[h]
  \centering
    \includegraphics[width=0.9\linewidth]{Fig4.png}
    \caption{Variance (measured by the standard deviation) in water velocity, shear stress, and width integrated shear stress for different time aggregations across the range of subglacial and subaerial channel shapes and friction factors.
      Shaded areas denote the minimum and maximum of the $100$ parameter combinations tested for each of the systems (Section~\ref{sect:imp}).
      Solid lines denote  mean values.
      Top row shows results from \alpine{} case, while bottom row show outputs from \icesheet{} case
      Time intervals shown with markers denote $15$\,\unit{min}, $1$\,\unit{hr}, $12$\,\unit{hr}, $1$\,\unit{d}, $5$\,\unit{d}, $10$\,\unit{d}, and $15$\,\unit{d} aggregation periods.
      Text in columns shows rank correlation between the variable and sediment discharge for the corresponding time period for the subglacial (right column) and subaerial (left column) channels.
    }
    \label{fig:multi_run}
  \end{figure}

\subsection{Relationship between parameterization  outputs and observed sediment discharge}
When comparing the output variables with suspended sediment discharge measurements, velocity, shear stress, and width-integrated shear stress from the subaerial system generally show better correlation with sediment discharge compared to subglacial outputs across the tested subglacial channel parameters (Figures~\ref{fig:Qw_vari} g--l and \ref{fig:multi_run}).
The rank correlation amongst the  subaerial parameterization's three output variables remains fairly constant despite the variable impact of channel shape and consistent scaling with water discharge (Figure \ref{fig:Qw_vari} and Section~\ref{sect:fluv}).
In subglacial channels, velocity and shear stress perform substantially worse in terms of rank correlation with sediment discharge compared to width-integrated shear stress (Figures~ \ref{fig:multi_run}).
In both \alpine{} and \icesheet{}, the  subglacial rank correlation approaches that of subaerial channels over longer time aggregation periods (Figures~ \ref{fig:multi_run} and \ref{fig:model_outs_1hr} to \ref{fig:model_outs_15day}).

\section{Discussion}
\subsection{Context of sediment capacity differences between the channel types}
\label{sect:dis_qsc}
Quantities that control sediment transport capacity, such as water velocity, shear stress, and width-integrated shear stress, vary differently in response to water discharge in subglacial channels compared to subaerial ones (Figure~\ref{fig:Qw_vari}).
In sediment transport relationships, such as in \citeA{meyer1948} or \citeA{engelund1967}, that evaluate sediment discharge capacity, shear stress is scaled to the $\frac{3}{2}$ or $\frac{5}{2}$ power, respectively.
The exponent greater than $1$ magnifies sediment discharge variability beyond the variable sediment transport parameters described above (Figure~\ref{fig:multi_run}).
Note that sediment transport relationships describe sediment flux per unit width so that the integrated capacity across channel occurs after application of the exponent, unlike the width-integrated shear stress evaluation here \cite<Equations\,\ref{eq:tautg} and \ref{eq:tautf}; c.f.>{tucker1997}.
Note also that both subglacial and subaerial parameterizations here assume that the distribution of velocity and shear stress is homogeneous across the channel bed, despite the limitations of this assumption \cite<e.g.>{yager2018}.

The rapid increases and decreases in subglacial sediment transport capacity may compound the variability in sediment transport already present in subaerial systems \cite<Figure~\ref{fig:Qw_vari}; >{williams1989,jerolmack2010}.
By applying the discharge - width scaling in Equation~\ref{eq:wcf} to Equation~\ref{eq:v}, by setting $S\, \propto \,Q^{\frac{1}{2}}$, the subaerial channel's width-integrated shear stress response is $\tau_{tf}\, \propto\,  Q$. 
For sufficiently short time periods,  channel width and wetted area are fixed in the subglacial system so that the response of width-integrated shear stress is $ \tau_{tg}\, \propto\,  Q^{2}$, with a larger exponent on $Q$ than in the subaerial case following \citeA{alley1997}.
In this case, water discharge fluctuates at a shorter timescale than the subglacial channel evolution \cite{gimbert2016}.
Note also that the subglacial width integrated shear stress response to water discharge will be greater than the subaerial's provided that  Equation~\ref{eq:wcf}'s exponent is below one.
However, when subglacial channel size lies in equilibrium with water discharge, the  width-integrated shear stress response is $\tau_{tg}\, \propto \, Q^{\frac{4}{5}}$, similar to subaerial channels.
It follows that, the stronger correlation between hydraulics and water discharge in \icesheet{} could result from being closer to this equilibrium between variations in the channels wetter area and water discharge compared to \alpine{}.
This state could result from the reduced diurnal discharge amplitude in \icesheet{} (Figures~\ref{fig:model_outs} and ~\ref{fig:Qw_vari}).
In both cases, hydraulics could trend toward this equilibrium regime when average discharge over $1-5$ days when variability in subglacial channels  approaches that of subaerial channels (Figure~\ref{fig:multi_run}).

In subglacial channels, peaks occur with the fastest rate of  water discharge increase relative to the channel's growth (Figure~\ref{fig:model_outs}).
In turn, the subglacial channels here contain feedbacks that limit sediment transport capacity as the subglacial channel grows with increasing water discharge capacity, effectively reducing sediment transport capacity with the stabilization of water discharge.
Similar feedback occurs in subaerial systems \cite{phillips2016}, albeit at longer timescales, as opposed to days here \cite{gimbert2016}.

First, these measurement stations lie $0.5$ to $2$\,\unit{km} downstream of the glacier terminus \cite{cowton2012,felix2022}, likely a long enough distance for the pressurized subglacial regime to transition to the open channel subaerial one.
The transition of subglacial to subaerial conditions may occur underneath glacier terminus \cite{perolo2018}, especially if closure rates of the ice channels are slow  \cite{rothlisberger1972}.
Furthermore,  if a transport-limited regime persists at the glacier margin, then sediment can be easily mobilized, and subaerial processes could be represented just a short distance downstream.

The consistent relationship between sediment discharge capacity and open channel flow conditions could be pronounced $\sim20$\,\unit{km} downstream of the Leverett site, where a strong correlation persists between sediment plume size and the river's water discharge into the Kangerlussuaq fjord \cite{mcgrath2010}.
In marine-terminating glacier catchments, for contrast, a less consistent relationship may occur between water discharge or melt extent and sediment plume size \cite{tedstone2012}.
Here, ocean water pressurizes subglacial channels at the ice front \cite<e.g.>{how2017}, so the minimal correlation could result from the inconsistent relationship between subglacial sediment transport capacity and water discharge (Figure~\ref{fig:Qw_vari}).
